\section{Conclusion and Future Work}

We studied label-free condition monitoring of a real transformer fleet from Dissolved Gas
Analysis and reported three results. (1) A severity-disentangled compositional
representation---the centred log-ratio of the combustible-gas composition---recovers fault
\emph{type} by construction, raising agreement with the Duval classes from ARI $0.16$ to $0.55$
with a simple linear projection; the neural (SD-CAE) and adversarial-independence variants are
documented negatives. (2) The field-event notes commonly used to validate such models are, in
this fleet, confounded by surveillance bias: a trivial sample count matches the gas chemistry
(AUC $0.76$ vs $0.74$, difference not significant) and a label-free score built from these
features loses its apparent forward validity once that count is controlled. (3) A label-free
health/risk index nonetheless outranks conventional severity baselines, and on a confound-free,
chemistry-defined target the gas trajectory weakly forecasts the onset of arcing.

\textbf{Limitations.} The dataset has no ground-truth fault labels, so every result is measured
against a rule (Duval) or a weak, biased note label; the fault-type agreement is moderate; and
the findings come from a single fleet. The methods are deliberately simple---a linear projection
in the right geometry and small multilayer perceptrons---rather than a new architecture.

\textbf{Future work.} The highest-leverage improvement is a better target---completing the
domain validation of our Thai-note translation (which already recovers 15 missed event notes)
or obtaining real failure/maintenance records---after which the risk and prognostic claims can
be strengthened. We
also see promise in confound-free, chemistry-defined prognostic targets, in survival/remaining-
useful-life modelling of gassing trends, and, following recent surveys \cite{ramarao2026review},
in physics-informed and digital-twin extensions. More broadly, we recommend that DGA studies
which validate against operator-logged events report the sample-count confound floor, as a
trivial surveillance signal can otherwise be mistaken for diagnostic skill.
